%*******************************************************
% Acknowledgments
%*******************************************************
\pdfbookmark[1]{Acknowledgments}{acknowledgments}

% \begin{flushright}{\slshape
%     We have seen that computer programming is an art, \\
%     because it applies accumulated knowledge to the world, \\
%     because it requires skill and ingenuity, and especially \\
%     because it produces objects of beauty.} \\ \medskip
%     --- \defcitealias{knuth:1974}{Donald E. Knuth}\citetalias{knuth:1974} \citep{knuth:1974}
% \end{flushright}
%
%

\bigskip

\begingroup
\let\clearpage\relax
\let\cleardoublepage\relax
\let\cleardoublepage\relax
\chapter*{Acknowledgments}
First and foremost I would like to thank Chris Yau for his wonderful supervision during my DPhil. 
He has taught me an incredible amount about statistics, genomics, and academic life in general, and in both our work and my plans in life he has encouraged me to go a level beyond that I would have before. On the rare occasion I'm convinced I've found a flaw in his thinking I'm swiftly proven wrong. I would also like to thank him for sending me to DisneyWorld.

I would like to thank Caleb Webber for his supervision on various projects and support and enthusiasm over the years. I am further indebted to Chris Ponting for his early supervision, but also his continued mentorship throughout my DPhil and advice in planning my future beyond these studies.

I'm also grateful to those with whom I've collaborated with scientifically both in Oxford and further afield. Davis McCarthy and Quin Wills on the development of Scater, Charmaine Lang on the Parkinson's project, Georg Holl{\"a}nder and Carlos Mayer on the Thymus project, and many others.

I am grateful to the Medical Research Council for funding my DPhil and also to St John's College Oxford for generous financial support for academic expenses and conference travel.

Many thanks to all my friends both in Oxford and further afield for the fun times we've had. I feel lucky to know such a great group of people.

I am deeply thankful to my parents for their love and support through 21 (!) continuous years of education and have enjoyed their visits to Oxford in the time I've been here.

Finally to Sally, for her support in listening to me complain about ``the statistics of single-cell biology'',
but really for the adventures we've gone on during this time that have made it all such fun.


\endgroup
